Diversification of Australian blind snakes

### 2022-10-31 manuscript.Rmd
update

# Introduction
<!-- COMMUNITY words - Reported, accepted, widely  -->
<!-- TRANSITION / FLOW words - Nonetheless, however, although, unless, because, and, but/ -->
<!-- 1. Why phylogeny is important to readers of the target journals? -->

The aridification on continental Australia over the past 20 million years resulted in a diversity of habitats and spurred multiple radiations of flora and fauna across the continent, many of which are endemic [@byrneBirth2008; @melvilleEvolutionary2011]. The arid zone makes up the majority of the Australian landscape and provides habitat to ~43% of all Australia’s terrestrial reptiles. Squamate reptiles (lizards and snakes) represent the largest vertebrate diversity in Australia with over 1000 species (~800 lizards and ~200 snakes), many of which have undergone rapid diversification in the last 10 – 6 myr [@brennanMass2017; @oliverPhylogenetic2017]. For example, some Australian squamate clades like dragons [@melvilleIntegrating2019], sphenomorphine skinks [@raboskyAnalysis2014], geckos [@ashmanDiversification2018; @brennanMass2017] and elapid snakes [@leeDiversification2016; @grundlerTrophic2014] show increased rates of diversification coincidentally with the onset of aridification. However, these studies did not account for heterogeneity of the arid zone [@pepperLife2021].

Roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus _Anilios_, which comprises 48 described species [@uetzQuarter2021]. All _Anilios_ species are found in Australia except for _A. eryncinus_, which is only found in Papua New Guinea. While morphologically conserved and difficult to distinguished, extensive phylogenetic studies and several efforts to revisit specimens revealed the true species diversity is underestimated [@marinHidden2013; @pyronSystematics2014; @ellisTyphlopid2019]. The diversification is reported to have originated from a single colonisation event from Asia between 19-39 mya [@vidalBlindsnake2010; @marinTracing2013], before the onset of aridification over the past 20 million years [@byrneBirth2008]. However, the study did not account for alternative biogeographic areas (e.g. WWF Ecoregions) and models, or uncertainty in phylogenetic relationships and divergence times. Despite representing a substantial portion of snake biodiversity there has been few investigations into the macroevolutionary dynamics of blind snakes. Recent phylogenomic tree and monophyly of the genus makes the Australian _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic ecological changes because they are geographically restricted to the continent and can be found in almost every biome [citefirstchapter?].

<!-- Environmental correlates of diversification -->
<!-- agamid lizards [@harmonTempo2003], passerine birds [@ricklefsGlobal2006], amphibians [@rollandContribution2019], rosid plants [@sunRecent2020]).  -->
A variety of ecological correlates have been associated with diversification processes of a clade. Among others, studies on the influence of climate (e.g. temperature, precipitation, and latitude) [@ricklefsGlobal2006; @mayhewBiodiversity2012; @condamineMacroevolutionary2013; @cantalapiedraDietary2014; @condamineClimate2019; @rollandContribution2019; @sunRecent2020]], elevation [@esquerreHow2019; @boschmanMountain2021], biological interactions [@ramirezAsynchronous2011; @druryEstimating2016; @brennanPhylogenomicsMonitorLizards2021], CO<sub>2</sub> [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017] and aridification [@raboskyExceptional2007; @pepperGenetic2011; @deepakAridification2018; @catulloAridification2014] have been documented. In Australia, shifts in diversification rates owing to changes in temperature and aridity has been reported in skinks [@raboskyExceptional2007], geckos [@fujitaDiversification2010; @pepperGenetic2011; @pepperPalaeoclimate2011; @oliverIndependent2014; @oliverContrasting2014], frogs [@catulloAridification2014; @vidal-garciaConvergent2015], saltbush [@kadereitMolecular2010], *Acacia* [@millerComparative2013], insects [@owenHow2017; @toussaintUnveiling2015], and other plants [reviewed in @crispHow2013].

There are several links between temperature and diversification rates of reptiles. First, warm temperature may increase resource availability, which in turn increase ecological niche in general [@clarkeClimate2006]. Second, spatial and temporal heterogeneity in temperature may lead to climatic niche divergence, such as local adaptations in different geographic areas [@pepperGeodiversity2013], and change in activity time of prey/predators [@santosTemperature2011; @lawsonLatitudinal2014]. Third, change in climate change (e.g. glaciation) may cause fragmentation of continuous habitats leading to isolated populations, which may lead to divergence [@oliverRocks2019]. Fourth, ectotherms are subjected to higher metabolic rates in warm climates (i.e. Kleiber's law), which may increase mutation rates and ultimately greater speciation rates [@allenKinetic2006; @makarievaMean2008]. Finally, aridification of Australia may offer new ecological niche for species to exploit, which have been shown to result in deep divergences in many arid-adapted [@byrneBirth2008; @pepperGenetic2011; @marinTracing2013].

The influence of temperature on diversification rates has been empirically tested with birth-death models [@condamineMacroevolutionary2013; @kergoatOpposite2018; @condamineClimate2019; @rollandContribution2019], these temperature data are reconstructed (interpolated) based on conversion of concentration of oxygen isotopes [@condamineMacroevolutionary2013]. However, the relationship between diversification rates and aridity are attributed to some proxy such as the the timing of aridification or global temperature rather than quantifying the extent of aridity itself [@dornStrange2014; @deepakAridification2018; @catulloAridification2014; @fujitaDiversification2010; @pepperGenetic2011; @pepperPalaeoclimate2011; @oliverIndependent2014; @oliverContrasting2014; @owenHow2017; @crispHow2013; @toussaintUnveiling2015]. As such, no study has been able to quantitatively assess aridity as a continuous variable in environment-dependent models and compare it to other ecological processes. Recent development in geographic-based methods such as `gen3sis` modeling engine [@hagenGen3sis2021] enables us to simulate climatic, geological, and ecological factors that are spatially and temporally specific to certain areas. In practice, these outputs from `gen3sis` provide paleotemperature and aridity data specific to Australia (instead of relying on global paleoclimate data) in the time span of interest (between the 45 mya and the present). These datasets will allow us to empirically test the influence of localised ecological processes on diversification rates over time.

In this study, we use the Australian radiation of blindsnakes as a system to investigate the diversification of morphologically conserved squamates in arid and non-arid zones by testing hypotheses about the processes that promote speciation. Applying the cutting edge localised geographic-based paleoenvironment reconstruction,

We (1) evaluated the model fit performance of localised versus globalised temperature on the diversification of Australian blindsnake radiation, (2) compared the model fit between different ecological processes on diversification rates, and (3) reconstructed the ancestral biome of Australian blindsnakes?

(4) Are there differences in the rates of speciation between arid and non-arid zones?

@@

To test whether diversification rates differs between arid and non-arid zones we used the R package GeoHiSSE [@caetanoHidden2018].

@@

- The emergence aridification has been hypothesised as a driver of diversification of many groups of reptiles and amphibians (CITE). 
- XX species today occupy arid zones exclusively and part of the range of YY species includes the arid zone as defined by WWF Bioregions. 
To evaluate whether diversification rates are associated with geographic ranges, we used a series of geographic range-dependent speciation and extinction models including GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] implemented in the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016]. 
- Both GeoSSE and GeoHiSSE allow only two geographic states, as such we categorised species that primarily occur in the arid zone as arid, and species that occur primarily outside of the arid zone as non-arid.

>>>>

# Introduction

<!-- Aridification  -->

The aridification on continental Australia over the past 20 million years resulted in a diversity of habitats and spurred multiple radiations of flora and fauna across the continent, many of which are endemic [@byrneBirth2008; @melvilleEvolutionary2011]. The arid zone makes up the majority of the Australian landscape and provides habitat to ~43% of all Australia’s terrestrial reptiles. Squamate reptiles (lizards and snakes) represent the largest vertebrate diversity in Australia with over 1000 species (~800 lizards and ~200 snakes), many of which have undergone rapid diversification in the last 10 – 6 myr [@brennanMass2017;@oliverPhylogenetic2017]. For example, some Australian squamate clades like dragons [@melvilleIntegrating2019], sphenomorphine skinks [@raboskyAnalysis2014], geckos [@ashmanDiversification2018;@brennanMass2017] and elapid snakes [@leeDiversification2016;@grundlerTrophic2014] show increased rates of diversification coincidentally with the onset of aridification. However, these studies did not account for heterogeneity of the arid zone [@pepperLife2021].

Roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus _Anilios_, which comprises **47** described species [@uetzQuarter2021]. All _Anilios_ species are found in Australia except for _A. eryncinus_, which is only found in Papua New Guinea. While morphologically conserved and difficult to distinguished, extensive phylogenetic studies and several efforts to revisit specimens revealed the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@ellisTyphlopid2019]. The diversification is reported to have originated from a single colonisation event by a non-arid adapted ancestor from Asia between 19-39 mya [@vidalBlindsnake2010;@marinTracing2013], before the onset of aridification over the past 20 million years [@byrneBirth2008]. 
However, the study did not account for alternative biogeographic areas (e.g. WWF Ecoregions) and models, or uncertainty in phylogenetic relationships and divergence times. Despite representing a substantial portion of snake biodiversity, there has been few investigations into the macroevolutionary dynamics of blind snakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic ecological changes because they are geographically restricted to the continent and can be found in almost every biome.

<!-- Environmental correlates of diversification -->
<!-- agamid lizards [@harmonTempo2003], passerine birds [@ricklefsGlobal2006], amphibians [@rollandContribution2019], rosid plants [@sunRecent2020]).  -->
A variety of ecological correlates have been associated with diversification processes of a clade. Among others, studies on the influence of climate (e.g. temperature, precipitation, and latitude) [@ricklefsGlobal2006;@mayhewBiodiversity2012;@condamineMacroevolutionary2013;@cantalapiedraDietary2014;@condamineClimate2019;@rollandContribution2019;@sunRecent2020], elevation [@esquerreHow2019;@boschmanMountain2021], biological interactions [@ramirezAsynchronous2011;@druryEstimating2016;@brennanPhylogenomicsMonitorLizards2021], CO<sub>2</sub> [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017] and aridification [@raboskyExceptional2007;@pepperGenetic2011;@deepakAridification2018;@catulloAridification2014] have been documented. In Australia, shifts in diversification rates owing to changes in temperature and aridity has been reported in skinks [@raboskyExceptional2007], geckos [@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014], frogs [@catulloAridification2014;@vidal-garciaConvergent2015], saltbush [@kadereitMolecular2010], *Acacia* [@millerComparative2013], insects [@owenHow2017;@toussaintUnveiling2015], and other plants (reviewed in @crispHow2013).

There are several links between temperature and diversification rates of reptiles. First, warm temperature may increase resource availability, which in turn increase ecological niche in general [@clarkeClimate2006]. Second, spatial and temporal heterogeneity in temperature may lead to climatic niche divergence, such as local adaptations in different geographic areas [@pepperGeodiversity2013], and change in activity time of prey/predators [@santosTemperature2011;@lawsonLatitudinal2014]. Third, change in climate (e.g. glaciation and aridification) may cause fragmentation of continuous habitats leading to isolated populations, which may lead to divergence [@oliverRocks2019]. Fourth, ectotherms are subjected to higher metabolic rates in warm climates (i.e. Kleiber's law), which may increase mutation rates and ultimately greater speciation rates [@allenKinetic2006;@makarievaMean2008]. Finally, aridification of Australia may offer new ecological niche for species to exploit, which have been shown to result in deep divergences in many arid-adapted [@byrneBirth2008;@pepperGenetic2011;@marinTracing2013].

The influence of temperature on diversification rates has been empirically tested with birth-death models [@condamineMacroevolutionary2013;@kergoatOpposite2018;@condamineClimate2019;@rollandContribution2019], these temperature data are reconstructed (interpolated) based on conversion of concentration of oxygen isotopes [@condamineMacroevolutionary2013]. However, the relationship between diversification rates and aridity are attributed to some proxy such as the the timing of aridification or global temperature rather than quantifying the extent of aridity itself [@dornStrange2014;@deepakAridification2018;@catulloAridification2014;@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014;@owenHow2017;@crispHow2013;@toussaintUnveiling2015]. As such, no study has been able to quantitatively assess aridity as a continuous variable in environment-dependent models and compare it to other ecological processes. Recent development in geographic-based methods such as `gen3sis` modeling engine [@hagenGen3sis2021] enables us to simulate climatic, geological, and ecological factors that are spatially and temporally specific to certain areas. In practice, these outputs from `gen3sis` provide paleotemperature and aridity data specific to Australia in the time span of interest, between the 45 MYA to present. These datasets will allow us to empirically test the influence of localised ecological processes on diversification rates over time.

In this study, we use the Australian radiation of blindsnakes as a system to investigate the diversification of morphologically conserved squamates by testing hypotheses about the processes that promote speciation.

<!--Why people should care-->
- Applying the cutting edge localised geographic-based paleoenvironment reconstruction,

We 
(1) evaluated the model fit performance of localised versus globalised temperature on the diversification of Australian blindsnake radiation, 
(2) compared the model fit between different ecological processes on diversification rates, and 
(3) reconstructed the ancestral biome of Australian blindsnakes? 
(4) Are there differences in the rates of speciation between arid and non-arid zones?

@@

- Aridification has been hypothesised as a driver of diversification of many groups of reptiles and amphibians (CITE). 
- XX species today occupy arid zones exclusively and part of the range of YY species includes the arid zone as defined by WWF Bioregions. 
- To evaluate whether broad patterns of diversification are associated with geographic range, we used a series of biogeographic-dependent speciation-extinction models 
including GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] implemented in the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016].

- GeoHiSSE differs from GeoSSE by incorporating hidden states, which accounts for heteregenous diversification caused by unsampled traits.

- These models only allow three states including two geographic states and one widespread state. As such we categorised species that primarily occur in the arid zone as arid, species that occur primarily outside of the arid zone as non-arid, and those that occupy both the arid and non-arid zone as widespread. We assumed uniform taxon sampling for 89% in arid areas, 71 in non-arid areas, and 83% in widespread areas. 
-

@@
